# Continuous Assessment 1

```{r}
ca1 <- read.csv("CA1.csv")
matrix_ca1 <- as.matrix(ca1)
```

------------------------------------------------------------------------

## Question 1

Compute and report the sample mean vectors for each of the five time periods.

```{r}
mean_matrix <- matrix(nrow=5, ncol=4)
for(i in 1:5){
  X <- matrix_ca1[matrix_ca1[,5] == i,]
  mean_matrix[i,] <- (1/nrow(X))*t(X[,-5])%*%matrix(1,nrow=nrow(X),ncol=1)
}
rownames(mean_matrix) <- c("MaxBreadth", "BaseHeight", "BasLength", "NasHeight", "TimePeriod")
columnnames(mean_matrix) <- c("4000 B.C", "3300 B.C.", "1850 B.C.", " 200 B.C.", " 150 A.D.")
mean_matrix
```

## Question 2

Provide a heat map of the correlation matrix for each time period and briefly interpret. Are there any noticeable changes over the time periods?

## Question 3

Calculate the angle between the deviation vectors for $X_1$ and $X_3$ in period 1. Explain why this value is to be expected by referring to the appropriate value from question 2. For a bonus mark, plot all the deviation vectors for period 1 across the first two observations.

## Question 4

Suppose researchers are interested in the quantity $Y_i = 3X_4 − X_1$ for time periods i = 1, . . . ,5. Use your answers from question 1 and an appropriate vector b to determine the sample means $\bar{y_1},..., \bar{y_5}$. Also give the covariance matrix of Y = \[Y1 Y2 Y3 Y4 Y5\]'
