# Continuous Assessment 1

```{r}
install.packages("ggcorrplot")
install.packages("gridExtra")
ca1 <- read.csv("CA1.csv")
X_total <- as.matrix(ca1)
```

------------------------------------------------------------------------

## Question 1

Compute and report the sample mean vectors for each of the five time periods.

```{r}
mean_matrix <- matrix(nrow=5, ncol=5)
for(i in 1:5){
  X_group <- X_total[X_total[,5] == i,]
  mean_matrix[i,] <- (1/nrow(X_group))*t(X_group)%*%matrix(1,nrow=nrow(X_group),ncol=1)
}
colnames(mean_matrix) <- c("MaxBreadth", "BaseHeight", "BasLength", "NasHeight", "TimePeriod")
rownames(mean_matrix) <- c("4000 B.C", "3300 B.C.", "1850 B.C.", " 200 B.C.", " 150 A.D.")
(mean_matrix <- t(mean_matrix))
```

------------------------------------------------------------------------

## Question 2

Provide a heat map of the correlation matrix for each time period

```{r}
library(ggcorrplot)
library(gridExtra)

for(i in 1:5){
  # X_bar for each time period
  X_bar <- matrix(mean_matrix[1:4,i])
  
  # X matrix for each time period
  X_group <- X_total[X_total[,5] == i,1:4]
  
  # Variance-Covariance Matrix
  S <- t(X_group) %*% X_group - nrow(X_group)*X_bar %*% t(X_bar)
  S <- S/(nrow(X_group) - 1)
  
  # Diagonals Matrix
  D <- diag(diag(S))
  
  # Correlation Matrix
  R <- solve(D)^(1/2) %*% S %*% solve(D)^(1/2)
  rownames(R) <- c("MaxBreadth", "BaseHeight", "BasLength", "NasHeight")
  colnames(R) <- c("MaxBreadth", "BaseHeight", "BasLength", "NasHeight")
  
  # Correlation Heat Map Plot
  print(ggcorrplot(R, lab = TRUE, lab_size = 3) + 
    ggtitle(paste("Correlation Matrix For Time Period:", i)) +
    theme(plot.title = element_text(hjust = 0.5, size=10), axis.text=element_text(size=8)))
}

```

And briefly interpret the heat map. Are there any noticeable changes over the time periods?

```{r}
# 1. The correlation between the MaxBreadth and NasHeight went from being positively correlated to each other in time period 1 to gradually becoming uncorrelated to each other by time period 5

# 2. The correlation between BaseHeight and BasLength went from being uncorrelated to each other in time period 1, to slowy become more and more positively correlated to each other by time period 5

# 3. The correlation between the MaxBreadth and BaseHeight went from being positively correlated in time period 1 to gradually became less and less uncorrelated by time period 5.
```

------------------------------------------------------------------------

## Question 3

Calculate the angle between the deviation vectors for $X_1$ and $X_3$ in period 1.

```{r}
# Deviation Vectors
X_center <- scale(X_total[X_total[,5] == 1,], center = T, scale = F) # same as di = xi - x_bar * 1
d1 <- matrix(X_center[,1])
d2 <- matrix(X_center[,3])

# cos(theta) = angle between deviation vectors X1 and X3
cos_theta <- (t(d1) %*% d2)/((sqrt(t(d1)%*%(d1)))*(sqrt(t(d2)%*%(d2))))
theta <- acos(cos_theta)
print(paste("Correlation:", cos_theta))
print(paste("Angle in degrees:", theta*180/pi))
print(paste("Angle in radians:", theta))
```

Explain why this value is to be expected by referring to the appropriate value from question 2.

```{r}
cor(X_total[X_total[,5] == 1,1:4])
# The correlation between X1 (MaxBreadth) and X3 (BasLength) is 0.01504250 which is the same as finding the cos(angle) between the two deviation vectors
cos(theta)
```

For a bonus mark, plot all the deviation vectors for period 1 across the first two observations.

```{r}
X_center <- scale(X_total[X_total[,5] == 1,1:4], center = T, scale = F) 

# Plot deviation Vectors
d1 <- matrix(X_center[1:2,1])
d2 <- matrix(X_center[1:2,2])
d3 <- matrix(X_center[1:2,3])
d4 <- matrix(X_center[1:2,4])

library(ggplot2)

# Base Plot Setup
plot(0,0, type = "n", xlim=c(-12,5), ylim=c(-8,0), xlab ="X Axis", ylab = "Y Axis", main = "Deviation Vectors") 

# Add Arrow 
arrows(0, 0, d1[1], d1[2], col = "blue")
arrows(0, 0, d2[1], d2[2], col = "red")
arrows(0, 0, d3[1], d3[2], col = "yellow")
arrows(0, 0, d4[1], d4[2], col = "green")

# Add labels near arrow tips
text(d1[1], d1[2], labels = "d1", pos = 4, col = "blue")
text(d2[1], d2[2], labels = "d2", pos = 4, col = "red")
text(d3[1], d3[2], labels = "d3", pos = 4, col = "yellow")
text(d4[1], d4[2], labels = "d4", pos = 4, col = "green")
```

------------------------------------------------------------------------

## Question 4

Suppose researchers are interested in the quantity $Y_i = 3X_4 − X_1$ for time periods i = 1, . . . ,5. Use your answers from question 1 and an appropriate vector b to determine the sample means $\bar{y_1},..., \bar{y_5}$. Also give the covariance matrix of Y = \[Y1 Y2 Y3 Y4 Y5\]'

```{r}
Y <- matrix(0, nrow=nrow(X_total[X_total[,5]==1,1:4]), ncol=5)
Y_Expectation <- matrix(0, nrow = 5,ncol=1)
b <- matrix(data = c(-1,0,0,3,0), nrow = 5, ncol = 1)

for(i in 1:5){
  # Yi = 3*X4-X1 so X*b_vector for each time period
  Y[,i] <- (X_total[X_total[,5]==i,]) %*% b
  
  # E(b'X) = E(b1*x1 + b2*x2 + b3*x3 + b4*x4 + b5*x5) = b1*E(x1) + b2*E(x2) + b3*E(x3) + b4*E(x4) + b5*E(x5)
  X_bar <- matrix(mean_matrix[,i])
  Y_Expectation[i] <- t(b) %*% X_bar
}

rownames(Y_Expectation) <- c("1", "2", "3", "4", "5")
colnames(Y) <- c("Y1", "Y2", "Y3", "Y4", "Y5")

print("Sample means yi for time periods i = 1,2,3,4,5:")
Y_Expectation

# Finding the covariance for Y
Y_bar <- (1/nrow(Y)) * t(Y) %*% matrix(1, nrow=30, ncol=1) # Same as Y_Expectation (checking)
S <- t(Y) %*% Y- nrow(Y)*Y_bar%*%t(Y_bar)
S <- S/(nrow(Y)-1)
S
```
